Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2017-09-11 article.tex
Implement changes of Reviewer 1 except for first issue

In this paper we present \pkg{dMod}, an \proglang{R} package on dynamic modeling and parameter estimation. The aim and core functionality of \pkg{dMod} is
\begin{enumerate}[(i)]
	\item facilitated set-up of dynamic models with automated \proglang{C} code generation for fast simulation of model predictions and model sensitivities,
	\item flexible set-up of general parameter transformations (explicit or implicit) and observation functions, allowing for the implementation of multiple experimental conditions similar to mixed-effects modeling,
	\item parameter estimation based on trust-region optimization of the negative log-likelihood, making use of the sensitivity equations of the dynamic system and of symbolic derivatives of the observation- and parameter transformation functions,
	\item identifiability and uncertainty analysis based on the profile-likelihood method to determine confidence intervals for parameters and predictions.
\end{enumerate}

@@

The consequent implementation of capabilities to generate and propagate derivatives on a compiled-code level distinguishes \pkg{dMod} from other modeling frameworks as discussed above. Most of the standard optimization routines implemented in \proglang{R} need derivative information. However, the computation of derivatives in the context of ODE models holds some pitfalls. The accuracy of sensitivities obtained from finite differences can be insufficient because the step control of the integrator presents an additional source of numeric inaccuracy. Even for numeric methods circumventing this problem, e.g.~complex-step derivatives \citep{squire1998using}, the use of sensitivity equations is still beneficial judging from the accuracy vs.~computational cost ratio. See \citep{raue2013lessons} for a comparison of methods.

@@

Parameter estimation is a common task in statistics. It describes the process of inferring parameter values or parameter ranges of a statistical model based on observed data. Over decades, appropriate estimators have been developed for different problem classes. The principle of maximum likelihood allows to derive an estimator which is especially suited for applications where the distribution of the measurement noise is known. This knowledge about the structure of the noise constitutes additional information that makes the maximum-likelihood estimator (MLE) \textit{efficient}, i.e.~from all unbiased estimators the MLE has the lowest variance. Other properties of maximum-likelihood estimation are \textit{consistency}, i.e.~the estimated parameter value approaches the true parameter value in the limit of infinite data-sample size, and \textit{asymptotic normality}, i.e.~in the limit of infinite data the MLE follows a multivariate normal distribution. See \citep{azzalini1996statistical} for an introduction to likelihood theory.

@@

Numerical optimization is a diverse field with as many algorithms as there are optimization problems around. For our application, derivative-based methods, and in particular the Newton method, play a major role.

@@

The Newton recursion, eq.~\eqref{eq:newton}, converges in one iteration if $l(\theta)$ is a quadratic form. However, $l(\theta)$ is only quadratic if the model $y(\theta)$ is linear in $\theta$ which is usually not the case for ODE models. On the other hand, each smooth function, no matter if quadratic or not, can be approximated by a quadratic function based on its Taylor series.
%Depending on the function's shape, the \textit{trust region}, i.e.~the region where $l(\theta)$ is well approximated, can become quite small. Accordingly, a Newton step pointing far beyond this region could be misleading and not helpful for optimization. This is the idea behind trust-region optimization \citep{wright1999numerical}.
Thus, $l(\theta)$ can be approximated by a quadratic function in at least a small region around $\theta$. The idea of trust-region optimization is to confine a Newton step to a region, the \textit{trust region}, where the quadratic approximation holds \citep{wright1999numerical}.
In each iteration, the trust-region radius is adjusted and the optimization problem restricted to the trust region is solved. For parameter estimation in ODE models this has the additional advantage that parameter changes are constricted and optimization steps into parameter regions where the ODE cannot be numerically solved any more  become less frequent. This make trust-region optimization the method of choice for ODEs.

@@

\subsection{Uncertainty analysis}

@@

A useful tool to derive confidence intervals beyond the scope of Fisher Information Matrix is the profile likelihood. Consider a parameter of interest, $\theta_i$, being one of the parameters in the vector $\theta$. Furthermore, let $c_{\tau}(\theta) = \theta_i - \tau = 0$ be a parameter constraint.
Then, extending the likelihood with the constraint via a Lagrange multiplier $\lambda$ yields
\begin{align}
	\begin{aligned}
		{\rm pl}_i:\quad \mathbb R & \longrightarrow \mathbb R\\
		\tau & \longmapsto \min_{(\theta, \lambda)} \big[ l(\theta) + \lambda c_{\tau}(\theta) \big]
	\end{aligned}
	\label{eq:pl}
\end{align}
which is called the profile likelihood of the $i^{\;\rm th}$ parameter. The path $\tau \mapsto (\hat\theta_{\tau}, \hat\lambda_{\tau})$ along the minimizing parameters is called the profile likelihood path. Hence, the profile likelihood ${\rm pl}_i$ returns the minimal log-likelihood under the constraint $\theta_i \stackrel{!}{=} \tau$. By construction, the inequality $D_i := {\rm pl}_i(\tau) - {\rm pl}_i(\hat \theta_i) \geq 0$ holds for all $\tau$, assuming equality at least for $\tau = \hat\theta_i$. Interpreting the unconstrained model as the null model, which we assume to be true, and that with $\theta_i$ fixed to $\tau$ as the alternative model, the value $D$ is twice the log of the likelihood-ratio between those. Hence, a likelihood-ratio test can be performed to accept or reject the alternative model at a given confidence level. To compute the corresponding threshold, we assume Wilks' theorem according to which the thresholds are the quantiles of the $\chi^2$ distribution with one degree of freedom. Consequently, the $68\%/90\%/95\%$-confidence intervals of $\theta_i$ are those values of $\tau$ for which $D_i(\tau) \leq 1$, $2.71$ and $3.84$, respectively.

@@

The profile likelihood approach can be applied to compute confidence intervals for model predictions, too, see \citep{kreutz2012likelihood, hass2016fast}.

@@

\pkg{dMod} makes use of derivatives wherever possible. It is one of the core functionalities of the \pkg{cOde} package upon which \pkg{dMod} is based to augment a system of ODEs by its sensitivity equations. If $r$ parameters are involved and the system consists of $n$ states, the number of equations grows as quick as $n^2 + nr$. Solving the equations can be considerably accelerated by utilizing compiled code. \pkg{dMod} provides the \code{odemodel} class. An object of class \code{odemodel} is generated from an \code{eqnlist} or \code{eqnvec}. In the background, \proglang{C} code for the ODE and the combined system of ODE and sensitivity equations is generated, written to the working directory and compiled. The \code{odemodel} object keeps track of the shared objects. It is the basis of a prediction function $x(t, p)$ generated by the \code{Xs()} command.

@@

A set of parameter values, forcings and events capture a certain condition in which we find the modeled system. Manipulating the system, single model parameters, forcings or events need to be changed to account for the manipulation. It is a typical approach in systems inference to systematically perturb small parts of a modeled system to reveal information about the processes.

@@

Objective functions can be added by the \code{"+"} operator, meaning the objective values, gradients and Hessians are accumulated by summation. Besides the standard function \code{normL2()}, \pkg{dMod} provides several other functions returning objective functions, e.g.~\code{constraintL2()} or \code{datapointL2()}. They allow to define parameter priors or treat data points as parameters, as shown in Section~\ref{sec:example}. Thus, a typical objective function used for parameter estimation could be
\begin{center}
	\code{obj <- normL2(data, g*x*(p1+p2+p3)) + constraintL2(mu, sigma)}.
\end{center}

@@

For the mathematical description of bile flow, a three-compartment differential equation model is used. TCA is pipetted into the buffer compartment (\code{TCA_buffer}) from where it is transported into the cells (\code{TCA_cell}). Intracellular TCA is exported back to the buffer or the canalicular compartment (\code{TCA_cana}). Finally, canalicular TCA flows back into the buffer compartment. These processes give rise to the differential equations and corresponding flowchart presented in Figure~\ref{fig:flowchart}.
\begin{figure}[ht]
	\centering
	\begin{minipage}{0.55\textwidth}
		\textbf{A}
		\begin{equation*}
			\begin{aligned}
				\frac{\rm d}{ {\rm d}t} \verb!TCA_buffer! &= \verb!-import * TCA_buffer!\\
				& \quad \verb! + export_sinus * TCA_cell!\\
				& \quad \verb! + reflux * TCA_cana!\\
				\frac{\rm d}{ {\rm d}t} \verb!TCA_cana! &= \verb!export_cana * TCA_cell!\\
				& \quad \verb! - reflux * TCA_cana!\\
				\frac{\rm d}{ {\rm d}t} \verb!TCA_cell! &= \verb!import * TCA_buffer!\\
				& \quad \verb! - export_sinus * TCA_cell!\\
				&\quad \verb! - export_cana * TCA_cell!.
			\end{aligned}
		\end{equation*}
	\end{minipage}
	\begin{minipage}{0.44\textwidth}
		\textbf{B}\\
	\includegraphics[width = \textwidth]{images/flowchart}
	\end{minipage}
	\caption{Differential equations and flowchart of the reaction network. Taurocholic acid,  TCA, is transported between three compartments by four different processes. (A) Assuming mass-action kinetics, the three dynamic states satisfy a set of coupled differential equations. (B) The equations are visualized in a flowchart. }
	\label{fig:flowchart}
\end{figure}

@@

The following code illustrates the implementation of the objective function and how it is used with the \code{trust()} optimizer from the \pkg{trust} package to obtain a model fit, shown in Figure~\ref{fig:myfit}.

@@

plotValues(myframe, value < 100)
plotPars(myframe, value < 100)
\end{CodeInput}
\end{CodeChunk}
Here, we have searched according to $\vec p_i = \vec p_0 + \Delta\vec p_i$ where $\vec p_0$ is the center, in our case \code{pouter}, $\Delta\vec p_i\sim N(0, \sigma^2)$ is a random parameter vector taken from a normal distribution, in our case $\sigma = 4$, and the index $i$ runs from 1 to \code{fits = 50}. The \code{mclapply()} command from the \pkg{parallel} package is used internally to run fits in parallel, here \code{cores = 4}. The result of \code{mstrust()} is a list of all returned values of \code{trust()}. To extract the final objective value, parameter values, convergence information and the number of iterations, \code{as.parframe()} is used. The multi-start approach identifies four local optima, see Figure~\ref{fig:mstrust}, which yield almost the same objective value, Figure~\ref{fig:mstrust}A.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure5}
	\caption{Result of multi-start fitting procedure. (A) Fits have been sorted by increasing objective value. Four optima were found with almost identical objective value. (B) The parameter values for different optima are shown in different colors. (C) Each local optimum corresponds to a different model prediction, shown in different colors. The observed states are pracitically undistinguishable although the internal states show different behavior.}
	\label{fig:mstrust}
\end{figure}
Despite the similar objective value, the optima are not close to each other in parameter space, as being illustrated by Figure~\ref{fig:mstrust}B, and lead to different predictions, Figure~\ref{fig:mstrust}C. 
Figure~\ref{fig:mstrust}B suggests that the two initial values \code{TCA_cell} and \code{TCA_cana} are connected in the sense that if one takes a large value, the other takes a small value and vice versa.
%From the initial value parameters \code{TCA_cell} and \code{TCA_cana} it seems that they are exchangeable. Where \code{TCA_cell} is high, \code{TCA_cana} is low and vice versa. 
This is not surprising because the observed cellular TCA amount is the sum of both. A new experiment needs to be designed to distinguish one situation from another.

@@

Alternatively, we want to implement the steady-state constraint by an \textit{implicit} parameter transformation, as opposed to the explicit transformation shown above. Let $\dot x = f(x, p)$ be our dynamic system. Then, under certain conditions, we find a function $g(p)$ such that $f\big( g(p), p\big) = 0$ for all $p$, i.e.~$x_S = g(p)$ is a steady state of $f$. The parameter transformation we want to generate is the function $p \mapsto \big(p, g(p)\big)$. Here, the set of outer parameters is the set of the reaction rates  $p$ whereas the set of inner parameters contains these rates and the corresponding steady states as initial value parameters. The root of $f$ must be determined numerically to which end \code{multiroot()} from the \pkg{rootSolve} package is used.

@@

More precisely, changing the data-point parameter \code{d1}, the model is quickly forced to
follow the new data point. This is apparent from the "validation" contribution
to the total objective value, Figure~\ref{fig:validation}A, since it remains small with respect to the "data"
contribution originating from all other data points. Forcing the model
prediction to deviate more than $0.01$ from the original one, the profile
exceeds the 95\% confidence threshold providing a confidence interval for the
prediction itself. By calculating
prediction profiles for several time points, a confidence band for the course of
\code{TCA\_cell} is constructed as shown in Figure~\ref{fig:validation}B. The
95\% confidence band is closed towards small and large amounts.

@@

\section{Extensions of dMod}
\label{sec:extensions}
Computer algebra and symbolic tools are not part of \proglang{R}'s core functionality. In this section we illustrate two symbolic tools that are shiped with \pkg{dMod},  dealing with structural non-identifiability and steady-state constraints. They are implemented in \proglang{Python} and are interfaced via the \pkg{rPython} package.

\subsection{Lie-group symmetry detection}
In Section \ref{chap_uncertainty}, profile likelihood computation showed the existence of both practically and structurally non-identifiable parameters. While practical non-identifiability arises from insufficient information in the data, structural non-identifiabilty is connected to Lie-group symmetries, i.e. transformations of the states and parameters \[\Psi: (x,\theta) \longmapsto (x^*, \theta^*)\] that preserve the model prediction of the observables: \[g(x^*, \theta^*) = g(x,\theta)\,.\] Based on \citep{merkt2015higher}, the \code{symmetryDetection()} command outputs a list of available symmetry transformations. For example, the code
\begin{CodeChunk}
\begin{CodeInput}
reactions <- NULL
reactions <- addReaction(reactions, "TCA_buffer", "TCA_cell", 
			 rate = "import_baso*TCA_buffer")
reactions <- addReaction(reactions, "TCA_cell", "TCA_buffer", 
			 rate = "export_sinus*TCA_cell")
reactions <- addReaction(reactions, "TCA_cell", "TCA_cana", 
			 rate = "export_cana*TCA_cell")
reactions <- addReaction(reactions, "TCA_cana", "TCA_buffer", 
			 rate = "reflux*TCA_cana")

@@

\subsection{Analytical steady-state constraints}
While for the the present model, the steady-state could be explicitly calculated by hand, this might be much more challenging for models with a large number of states and parameters. For many of these models, the \code{steadyStates()} command outputs an analytical steady-state solution that can be incorporated in the model as an additional parameter transformation. Based on \citep{rosenblatt2016customized}, the steady-state constraint is solved for a combination of state variables and kinetic parameters while positivity of the solution is ensured. As the paper states, the approach outperforms common methods of steady-state implementation with respect to reliability and performance of the optimization process. For our example, the code reads
\begin{CodeChunk}
\begin{CodeInput}
steadyStates(reactions, file = "SS.Rds)
\end{CodeInput}
\end{CodeChunk}
yielding the output

>>>>

In this paper we present \pkg{dMod}, an \proglang{R} package on dynamic modeling and parameter estimation. 
The package is grown over the last years and reflects many developments and lessons learned from our research projects with time-resolved experimental data from Systems Biology. They form the basis of the development of \pkg{dMod} and are mostly referenced in the following paragraphs.\\
The aim and core functionality of \pkg{dMod} is
\begin{enumerate}[(i)]
	\item facilitated set-up of dynamic models with automated \proglang{C} code generation for fast simulation of model predictions and model sensitivities,
	\item flexible set-up of general parameter transformations (explicit or implicit) and observation functions, allowing for the implementation of multiple experimental conditions similar to mixed-effects modeling,
	\item parameter estimation based on trust-region optimization of the negative log-likelihood, making use of the sensitivity equations of the dynamic system and of symbolic derivatives of the observation- and parameter transformation functions,
	\item identifiability and uncertainty analysis based on the profile-likelihood method to determine confidence intervals for parameters and predictions.
\end{enumerate}

@@

The consequent implementation of capabilities to generate and propagate derivatives on a compiled-code level distinguishes \pkg{dMod} from other modeling frameworks as discussed above. Most of the standard optimization routines implemented in \proglang{R} need derivative information. However, the computation of derivatives in the context of ODE models holds some pitfalls. The accuracy of sensitivities obtained from finite differences can be insufficient because the step control of the integrator presents an additional source of numeric inaccuracy. Even for numeric methods circumventing this problem, e.g.~complex-step derivatives \citep{squire1998using}, the use of sensitivity equations is still beneficial judging from the accuracy vs.~computational cost ratio. See \cite{raue2013lessons} for a comparison of methods.

@@

Parameter estimation is a common task in statistics. It describes the process of inferring parameter values or parameter ranges of a statistical model based on observed data. Over decades, appropriate estimators have been developed for different problem classes. The principle of maximum likelihood allows to derive an estimator which is especially suited for applications where the distribution of the measurement noise is known. This knowledge about the structure of the noise constitutes additional information that makes the maximum-likelihood estimator (MLE) \textit{efficient}, i.e.~from all unbiased estimators the MLE has the lowest variance. Other properties of maximum-likelihood estimation are \textit{consistency}, i.e.~the estimated parameter value approaches the true parameter value in the limit of infinite data-sample size, and \textit{asymptotic normality}, i.e.~in the limit of infinite data the MLE follows a multivariate normal distribution. See \cite{azzalini1996statistical} for an introduction to likelihood theory.

@@

Numerical optimization is a diverse field with as many algorithms as there are optimization problems around. The \pkg{dMod} package supports derivative-based methods, and in particular the Newton method.

@@

The Newton recursion, eq.~\eqref{eq:newton}, converges in one iteration if $l(\theta)$ is a quadratic form. However, $l(\theta)$ is only quadratic if the model $y(\theta)$ is linear in $\theta$ which is usually not the case for ODE models. On the other hand, each smooth function, no matter if quadratic or not, can be approximated by a quadratic function based on its Taylor series.
%Depending on the function's shape, the \textit{trust region}, i.e.~the region where $l(\theta)$ is well approximated, can become quite small. Accordingly, a Newton step pointing far beyond this region could be misleading and not helpful for optimization. This is the idea behind trust-region optimization \citep{wright1999numerical}.
Thus, $l(\theta)$ can be approximated by a quadratic function in at least a small region around $\theta$. The idea of trust-region optimization is to confine a Newton step to a region, the \textit{trust region}, where the quadratic approximation holds \citep{wright1999numerical}.
In each iteration, the trust-region radius is adjusted and the optimization problem restricted to the trust region is solved. For parameter estimation in ODE models this has the additional advantage that parameter changes are constricted and optimization steps into parameter regions where the ODE cannot be numerically solved any more  become less frequent. This makes trust-region optimization the method of choice for \pkg{dMod}.

@@

\subsection{Parameter uncertainty analysis}

@@

A useful tool to derive confidence intervals beyond the scope of Fisher Information Matrix is the profile likelihood \citep{venzon1988method, murphy2000profile}. Consider a parameter of interest, $\theta_i$, being one of the parameters in the vector $\theta$. Furthermore, let $c_{\tau}(\theta) = \theta_i - \tau = 0$ be a parameter constraint.
Then, extending the likelihood with the constraint via a Lagrange multiplier $\lambda$ yields
\begin{align}
	\begin{aligned}
		{\rm pl}_i:\quad \mathbb R & \longrightarrow \mathbb R\\
		\tau & \longmapsto \min_{(\theta, \lambda)} \big[ l(\theta) + \lambda c_{\tau}(\theta) \big]
	\end{aligned}
	\label{eq:pl}
\end{align}
which is called the profile likelihood of the $i^{\;\rm th}$ parameter. The path $\tau \mapsto (\hat\theta_{\tau}, \hat\lambda_{\tau})$ along the minimizing parameters is called the profile likelihood path. Hence, the profile likelihood ${\rm pl}_i$ returns the minimal log-likelihood under the constraint $\theta_i \stackrel{!}{=} \tau$. By construction, the inequality $D_i := {\rm pl}_i(\tau) - {\rm pl}_i(\hat \theta_i) \geq 0$ holds for all $\tau$, assuming equality at least for $\tau = \hat\theta_i$. Interpreting the unconstrained model as the null model, which we assume to be true, and that with $\theta_i$ fixed to $\tau$ as the alternative model, the value $D$ is twice the log of the likelihood-ratio between those. Hence, a likelihood-ratio test can be performed to accept or reject the alternative model at a given confidence level. To compute the corresponding threshold, we assume Wilks' theorem according to which the thresholds are the quantiles of the $\chi^2$ distribution with one degree of freedom. Consequently, the $68\%/90\%/95\%$-confidence intervals of $\theta_i$ are those values of $\tau$ for which $D_i(\tau) \leq 1$, $2.71$ and $3.84$, respectively.

@@

The profile likelihood approach can be applied to compute confidence intervals for model predictions, too, see \cite{kreutz2012likelihood, hass2016fast}.

@@

\pkg{dMod} makes use of derivatives wherever possible. It is one of the core functionalities of the \pkg{cOde} package \citep{cOde} upon which \pkg{dMod} is based to augment a system of ODEs by its sensitivity equations. If $r$ parameters are involved and the system consists of $n$ states, the number of equations grows as quick as $n^2 + nr$. Solving the equations can be considerably accelerated by utilizing compiled code. \pkg{dMod} provides the \code{odemodel} class. An object of class \code{odemodel} is generated from an \code{eqnlist} or \code{eqnvec}. In the background, \proglang{C} code for the ODE and the combined system of ODE and sensitivity equations is generated, written to the working directory and compiled. The \code{odemodel} object keeps track of the shared objects. It is the basis of a prediction function $x(t, p)$ generated by the \code{Xs()} command.

@@

A set of parameter values, forcings and events captures a certain condition in which we find the modeled system. Manipulating the system, single model parameters, forcings or events need to be changed to account for the manipulation. It is a typical approach in systems inference to systematically perturb small parts of a modeled system to reveal information about the processes.

@@

Objective functions can be added by the \code{"+"} operator, meaning the objective values, gradients and Hessians are accumulated by summation. Besides the standard function \code{normL2()}, \pkg{dMod} provides several other functions returning objective functions, e.g.~\code{constraintL2()} or \code{datapointL2()}. They allow to define quadratic parameter priors or treat data points as parameters, respectively, as shown in Section~\ref{sec:example}. Thus, a typical objective function used for parameter estimation could be
\begin{center}
	\code{obj <- normL2(data, g*x*(p1+p2+p3)) + constraintL2(mu, sigma)}.
\end{center}

@@

For the mathematical description of bile flow, a three-compartment differential equation model is used. TCA is pipetted into the buffer compartment (\code{TCA_buffer}) from where it is transported into the cells (\code{TCA_cell}). Intracellular TCA is exported back to the buffer or the canalicular compartment (\code{TCA_cana}). Finally, canalicular TCA flows back into the buffer compartment. These processes give rise to the differential equations and corresponding flowchart presented in Figure~\ref{fig:flowchart}.
\begin{figure}[ht]
	\centering
	\begin{minipage}{0.55\textwidth}
		\textbf{A}
		\begin{equation*}
			\begin{aligned}
				\frac{\texttt d}{ \texttt{dt} } \verb!TCA_buffer! &\texttt{ = } \verb!-import * TCA_buffer!\\
				& \quad \verb! + export_sinus * TCA_cell!\\
				& \quad \verb! + reflux * TCA_cana!\\
				\frac{\texttt d}{ \texttt{dt}} \verb!TCA_cana! &\texttt{ = } \verb!export_cana * TCA_cell!\\
				& \quad \verb! - reflux * TCA_cana!\\
				\frac{\texttt d}{ \texttt{dt}} \verb!TCA_cell! &\texttt{ = } \verb!import * TCA_buffer!\\
				& \quad \verb! - export_sinus * TCA_cell!\\
				&\quad \verb! - export_cana * TCA_cell!.
			\end{aligned}
		\end{equation*}
	\end{minipage}
	\begin{minipage}{0.44\textwidth}
		\textbf{B}\\
	\includegraphics[width = \textwidth]{images/flowchart}
	\end{minipage}
	\caption{Differential equations and flowchart of the reaction network. Taurocholic acid,  TCA, is transported between three compartments by four different processes. (A) Assuming mass-action kinetics, the three dynamic states satisfy a set of coupled differential equations. (B) The equations are visualized in a flowchart. }
	\label{fig:flowchart}
\end{figure}

@@

The following code illustrates the implementation of the objective function and how it is used with the \code{trust()} optimizer from the \pkg{trust} package \citep{trust} to obtain a model fit, shown in Figure~\ref{fig:myfit}.

@@

plotValues(myframe, value < 100)
plotPars(myframe, value < 100)
\end{CodeInput}
\end{CodeChunk}
Here, we have searched according to $\vec p_i = \vec p_0 + \Delta\vec p_i$ where $\vec p_0$ is the center, in our case \code{pouter}, $\Delta\vec p_i\sim N(0, \sigma^2)$ is a random parameter vector taken from a normal distribution, in our case $\sigma = 4$, and the index $i$ runs from 1 to \code{fits = 50}. The \code{mclapply()} command from the standard \pkg{parallel} package \citep{parallel} is used internally to run fits in parallel, here \code{cores = 4}. The result of \code{mstrust()} is a list of all returned values of \code{trust()}. To extract the final objective value, parameter values, convergence information and the number of iterations, \code{as.parframe()} is used. The multi-start approach identifies four local optima, see Figure~\ref{fig:mstrust}, which yield almost the same objective value, Figure~\ref{fig:mstrust}A.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure5}
	\caption{Result of multi-start fitting procedure. (A) Fits have been sorted by increasing objective value. Four optima were found with almost identical objective value. (B) The parameter values for different optima are shown in different colors. (C) Each local optimum corresponds to a different model prediction, shown in different colors. The observed states are pracitically undistinguishable although the internal states show different behavior.}
	\label{fig:mstrust}
\end{figure}
Despite the similar objective value, the optima are not close to each other in parameter space, as being illustrated by Figure~\ref{fig:mstrust}B, and lead to different predictions, Figure~\ref{fig:mstrust}C. 
Figure~\ref{fig:mstrust}B suggests that the two initial values \code{TCA_cell} and \code{TCA_cana} are connected in the sense that if one takes a large value, the other takes a small value and vice versa.
%From the initial value parameters \code{TCA_cell} and \code{TCA_cana} it seems that they are exchangeable. Where \code{TCA_cell} is high, \code{TCA_cana} is low and vice versa. 
This is not surprising because the observed cellular TCA amount is the sum of both. A new experiment needs to be designed to distinguish one situation from another.

@@

Alternatively, we want to implement the steady-state constraint by an \textit{implicit} parameter transformation, as opposed to the explicit transformation shown above. Let $\dot x = f(x, p)$ be our dynamic system. Then, under certain conditions, we find a function $g(p)$ such that $f\big( g(p), p\big) = 0$ for all $p$, i.e.~$x_S = g(p)$ is a steady state of $f$. The parameter transformation we want to generate is the function $p \mapsto \big(p, g(p)\big)$. Here, the set of outer parameters is the set of the reaction rates  $p$ whereas the set of inner parameters contains these rates and the corresponding steady states as initial value parameters. The root of $f$ must be determined numerically to which end \code{multiroot()} from the \pkg{rootSolve} package \citep{rootSolve} is used .

@@

More precisely, changing the data-point parameter \code{d1}, the model is quickly forced to
match the new data point. This is apparent from the least squares contribution entitled "validation" of the virtual data point, returned by \code{obj.validation}. 
The contribution is shown as dashed line in Figure~\ref{fig:validation}A. It remains small at the expense of a larger deviation from original data points, indicated by the "data" contribution. 
Forcing \code{d1} and thereby the model
prediction to deviate more than $0.01$ from the original value, the profile
exceeds the 95\% confidence threshold providing a confidence interval for the
prediction itself. By calculating
prediction profiles for several time points, a confidence band for the course of
\code{TCA\_cell} is constructed as shown in Figure~\ref{fig:validation}B. The
95\% confidence band is closed towards small and large amounts.

@@

\section{Extensions of dMod}
\label{sec:extensions}
Computer algebra and symbolic tools are not part of \proglang{R}'s core functionality. In this section we illustrate two symbolic tools that are shiped with \pkg{dMod},  dealing with structural non-identifiability and steady-state constraints. They are implemented in \proglang{Python} and are interfaced via the \pkg{rPython} package \citep{rPython}.

\subsection{Lie-group symmetry detection}
In Section \ref{chap_uncertainty}, profile likelihood computation showed the existence of both practically and structurally non-identifiable parameters. While practical non-identifiability arises from insufficient information in the data, structural non-identifiabilty is connected to Lie-group symmetries, i.e. transformations of the states and parameters \[\Psi: (x,\theta) \longmapsto (x^*, \theta^*)\] that preserve the model prediction of the observables: \[g(x^*, \theta^*) = g(x,\theta)\,.\] Based on \cite{merkt2015higher}, the \code{symmetryDetection()} command outputs a list of available symmetry transformations. For example, the code
\begin{CodeChunk}
\begin{CodeInput}
reactions <- NULL
reactions <- addReaction(reactions, "TCA_buffer", "TCA_cell", 
			 rate = "import_baso*TCA_buffer")
reactions <- addReaction(reactions, "TCA_cell", "TCA_buffer", 
			 rate = "export_sinus*TCA_cell")
reactions <- addReaction(reactions, "TCA_cell", "TCA_cana", 
			 rate = "export_cana*TCA_cell")
reactions <- addReaction(reactions, "TCA_cana", "TCA_buffer", 
			 rate = "reflux*TCA_cana")

@@

\subsection{Analytical steady-state constraints}
While for the the present model, the steady-state could be explicitly calculated by hand, this might be much more challenging for models with a large number of states and parameters. For many of these models, the \code{steadyStates()} command outputs an analytical steady-state solution that can be incorporated in the model as an additional parameter transformation. Based on \cite{rosenblatt2016customized}, the steady-state constraint is solved for a combination of state variables and kinetic parameters while positivity of the solution is ensured. As the paper states, the approach outperforms common methods of steady-state implementation with respect to reliability and performance of the optimization process. For our example, the code reads
\begin{CodeChunk}
\begin{CodeInput}
steadyStates(reactions, file = "SS.Rds")
\end{CodeInput}
\end{CodeChunk}
yielding the output
